---
title: "EDA_CDCFluView"
author: "Niat Kashsay, Celina Velazquez, Nancy Walker"
format: 
  pdf:
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{float}      % enables [H]
        \usepackage{placeins}   % provides \FloatBarrier
        \floatplacement{figure}{H}
        \floatplacement{table}{H}
execute:
  echo: true
  results: asis                    # show results inline
  warning: false
  message: false
---

# EDA of the CDCFluView Dataset 

The dataset researched comes from the CDC FLUVIEW.

https://gis.cdc.gov/grasp/fluview/fluportaldashboard.html

Data was accessed via the cdcfluview R package (v 0.x.x) on October 31, 2025, using the Influenza-Like Illness Network (ILINet) endpoint. Analyses reflect weekly ILI values available up to that date. 

Weekly ILI data for California were obtained from CDC ILINet (1997–present). For modeling, data from 2010 to 2025 were used to ensure consistent reporting and reliable seasonal structure. This dataset displays weekly influenza data. Influenza incidence can inform public health decision-making, such as surveillance, hospital capacity, and vaccination strategies. Although the CDC FLUVIEW provides influenza data nationwide, tailoring a project to Region 9 (California, Arizona, Hawaii, Nevada, and Pacific territories) provides stable weekly data that models well and yields a high-quality forecasting model. 

## Import libraries

```{r}
#| message: false
library(tidyverse)
library(fpp3)
library(janitor)   # for clean_names()
library(MMWRweek) 
library(gt)
library(corrplot)
```

## Read the Data 

```{r}
ili_raw <- read_csv("data/raw/ILINet.csv", 
                   skip =1, 
                   na = c("", "NA"),
                   show_col_types = FALSE)

# Make column names easy to use 
# For example, % Weighted ILI to percent_weighted_ili
ili <- ili_raw |>
  clean_names() |>
  # keep only Region 9 
  # (California, Arizona, Hawaii, Nevada, and pacific territories) 
  filter(region_type == "HHS Regions", region == "Region 9") |>
  # Create a simple sequence for weeks (1:nrow)
  mutate(
    # Create a true calendar date for the CDC week ending (MMWR day 7 = Sunday)
    week_end = MMWRweek2Date(MMWRyear = year, MMWRweek = week, MMWRday = 7)
  ) |>
  filter(!is.na(week_end)) |>
  as_tsibble(index = week_end)

# head(ili) 
```

## EDA

```{r}
# View Structure 
str(ili) 
```

Time index intervals are weekly (interval = 7D, regular = TRUE). Date is set as a tsibble index. Only Region 9 is selected. 

```{r}
# Check for null values 
anyNA(ili)
# Count null values 
sum(is.na(ili))
# Count by column 
colSums(is.na(ili))
```

Only age_25_64 has missing values and all entries are missing. All other columns have no missing values. This variable can be excluded from the dataset.

```{r}
# drop age_25_64 column form dataset 
ili <- ili |> select(-age_25_64)
# Rename age_65 to age_65_plus for clarity
ili <- ili |> 
  rename(age_65_plus = age_65)
```

```{r}
# View Summary 
summary(ili)
```

There are 781 observations, which is about 15 years of data from 2010 to 2025. The frequency of the data is weekly. Both short-term fluctuations and long-term seasonality can be analysed. 

*Target Variable* 

The target Time series variable is percent weight ili, representing the proportion of patient visits due to influenza-like illness (ILI) in Region 9, weighted by provider volume.

*Other Variables* 

Predictors (potentially explanatory) — e.g., age-group counts, number of providers, total patients.

Metadata or identifiers — e.g., year, week, region.

Alternative measures — e.g., percent_unweighted_ili (a variant of the same phenomenon).

### Trend and Seasonality 

```{r}
# Line plot of % Weighted ILI (overall trend)
autoplot(ili, percent_weighted_ili) +
  labs(title = "Region 9: % Weighted ILI (Weekly, 2010–2025)",
       y = "% ILI", x = "Week Ending Date")

ili |> 
  ggplot(aes(week_end, percent_unweighted_ili)) + 
  geom_line() + 
  # Apply the generalized additive model "gam"
  # Draw a flexible smooth curve throughout the data usign cubic splines 
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), se= FALSE) + 
  labs(title= "Weekly Weighted ILI (%) _ HHS Region 9", 
          subtitle = "Smoothed Trend Using Generalized Additive Model",
          x = "Week Ending Date",
          y = "Weighted ILI (%)")

```
```{r}
# Seasonal pattern across years
gg_season(ili, percent_weighted_ili, period = "year") +
  labs(title = "Seasonality by Year: % Weighted ILI",
       y = "% ILI", x = "Week of Year")

# decomposition plot 
ili |> 
  model (stl = STL(percent_weighted_ili)) |> 
  components() |> 
  autoplot()
```

### Distribution and Outliers 

```{r}
# Boxplot 
boxplot(ili$percent_weighted_ili,
        main = "Boxplot of percent_weighted_ili")

# Boxplot by year
ili |>
  mutate(year = factor(year(week_end))) |>
  ggplot(aes(year, percent_weighted_ili)) +
  geom_boxplot(outlier.alpha = 0.4, fill = "lightblue") +
  labs(title = "% Weighted ILI by Year", x = "Year", y = "% ILI")

# Histogram 
hist(ili$percent_weighted_ili,
     main="Histogram of percent_weighted_ili")

# Histogram of %ILI values
ili |> 
  ggplot(aes(percent_weighted_ili)) +
  geom_histogram(bins = 40, fill = "skyblue", color = "white") +
  labs(title = "Distribution of % Weighted ILI", x = "% ILI")

```
### Related Series and Context

**Weighted vs Unweighted Comparison** 
```{r}
# Compare weighted vs unweighted variables 
ili |>
  select(week_end, percent_weighted_ili, percent_unweighted_ili) |>
  pivot_longer(-week_end, names_to = "metric", values_to = "value") |>
  ggplot(aes(week_end, value, color = metric)) +
  geom_line() +
  labs(title = "Weighted vs Unweighted % ILI", x = "Week", y = "% ILI") 
```

```{r}
# Reporting coverage over time
ili |>
  pivot_longer(c(num_of_providers, total_patients),
               names_to = "measure", values_to = "value") |>
  ggplot(aes(week_end, value)) +
  geom_line() +
  facet_wrap(~ measure, scales = "free_y", ncol = 1) +
  labs(title = "Reporting Context: Providers & Patients", x = "Week")
```

### Age Group Insight

```{r}
age_cols <- c("age_0_4","age_5_24","age_25_49","age_50_64","age_65_plus")

ili |>
  select(week_end, all_of(age_cols)) |>
  pivot_longer(-week_end, names_to = "age_group", values_to = "count") |>
  ggplot(aes(week_end, count)) +
  geom_line() +
  facet_wrap(~ age_group, scales = "free_y", ncol = 2) +
  labs(title = "Weekly ILI Counts by Age Group", x = "Week End", y = "Count")
```
###Correlation Checks

```{r}
# make it a plain tibble, keep only numeric columns
ili_num <- ili |>
  as_tibble() |>
  select(where(is.numeric))

# Simple correlation matrix (numeric vars only)
ili_cor <- cor(ili_num, use= "pairwise.complete.obs")

corrplot(ili_cor, method = "color", type = "lower",
         tl.col = "black", tl.srt = 45)

# Focus on correlation with target variable
sort(ili_cor["percent_weighted_ili", ], decreasing = TRUE)
```

**Autocorrelation and Stationarity** 
```{r}
# Autocorrelation Function (ACF)
ACF(ili, percent_weighted_ili) |> autoplot() +
  labs(title = "Autocorrelation (ACF): % Weighted ILI")

# Partial Autocorrelation Function (PACF)
PACF(ili, percent_weighted_ili) |> autoplot() +
  labs(title = "Partial Autocorrelation (PACF): % Weighted ILI")
```

###Heatmap (Epidemiological View)
```{r}
ili |>
  mutate(year = year(week_end),
         week = week(week_end)) |>
  ggplot(aes(week, year, fill = percent_weighted_ili)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(title = "Heatmap of % Weighted ILI (Week vs Year)",
       x = "MMWR Week", y = "Year", fill = "% ILI")
```

## Refrences

OpenAI. (2025). ChatGPT (November 12 version) [LargeLanguage Model]